SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2012-12-29 paper/paper.tex
paper: updates.

class Arbitrary a => SubTypes a where
  size    :: a -> Size
  index   :: a -> Idx -> Maybe SubVal
  replace :: a -> Idx -> SubVal -> a
\end{code}
%
\noindent
\ttp{SubTypes} has three methods.  The \ttp{size} method returns the size of a
value, \ttp{index} returns a sub-value at an breadth-first index in a value, and
\ttp{replace} replaces a sub-value at a particular focus.  If the index is
out-of-bounds (too large or negative), \ttp{index} return \ttp{Nothing}.  The
method \ttp{replace} returns the original value for an out-of-bounds index or an
incorrectly-typed replacement value.  The \ttp{SubVal} type is an existential
data type, used as a generic container for sub-values from a counterexample.
%
\begin{code}
data SubVal =
  forall a. SubTypes a => SubVal a
\end{code}
%
\noindent
For example, consider a binary tree type
%
\begin{code}
data Tree = L | B Tree Tree
\end{code}
%
\noindent
and the value \ttp{tree}, labeled with the indexes of its sub-values
%
\begin{code}
tree = B\sub{0} (B\sub{1} L\sub{3}
             (B\sub{4} L\sub{6} L\sub{8}))
          (B\sub{2} L\sub{5} L\sub{7})
\end{code}
%
\noindent
Here are example applications of \ttp{SubTypes}; in the following, we show
the indexes with respect to the \ttp{tree} value:
%
\begin{code}
size tree = 9

index tree 4 = (Just . SubVal) (B\sub{4} L\sub{6} L\sub{8})

index tree 12 = Nothing

replace tree 2 (SubVal L) =
  B\sub{0} (B\sub{1} L\sub{3}
        (B\sub{4} L\sub{6} L\sub{8}))
     L
\end{code}
%
\noindent
We will sometimes refer to unwrapped value returned by \ttp{index a i} as the
\emph{$i$th sub-value of $a$}, so for example, \ttp{B\sub{4} L\sub{6} L\sub{8}}
is the 4th sub-value of \ttp{tree}.

\begin{figure}[ht]
  \lstset{caption={}
%         ,xleftmargin=5ex,firstnumber=0,numbers=left
         }
  \begin{code}
reduce :: SubTypes a
       => a -> (a -> Property) -> IO a
reduce cex prop = reduce' 1
  where
  reduce' idx =
    case index cex idx of
      Nothing -> return cex
      Just v  -> do
        vs <- newVals v (getSize v) scMaxFailure
        case test cex idx vs prop of
          Nothing   -> reduce' (idx+1)
          Just cex' -> reduce cex' prop

getSize :: SubVal -> Size
getSize (SubVal a) = size a

newVals :: SubVal -> Size -> Int -> IO [SubVal]
newVals (SubVal a) sz tries =
  sequence (replicate tries s)
  where
  s  =     sizedArbitrary sz a
       >>= return . SubVal

@@

\ttp{sizedArbitrary} returns a value inside the \ttp{IO} monad since it depends
on a random seed for generation (we could pass in the generator seed, but for
simplicity, we assume one is generated locally).  The constant \ttp{scMaxFailure} is
the maximum number of tries to discover a new counterexample by replacing
\ttp{v} in \ttp{cex} and testing it.  The result of \ttp{pass prop cex'} for
%
\begin{code}
pass :: (a -> Property) -> a -> Bool
\end{code}
%
\noindent
holds if \ttp{cex'} satisfies the property \ttp{prop}.  If no failure is found,
we move to the next sub-value of \ttp{cex} and continue.  However, if a new
smaller counterexample \ttp{cex'} is found, we start a new breadth-first
traversal of \ttp{cex'}, attempting to shrink it further.

@@

  test' v idx = do
    vs <- newVals v scMaxFailure
    case test cex idx vs prop of
      Nothing   -> reduce' (idx+1)
      Just cex' -> reduce0 cex' prop

@@

SmartCheck's implementation contains flags to allow the user to customize its
behavior.  Three flags that are relevant to the reduction algorithm are the
following:
%
\begin{code}
scMaxFailure :: Int
scMaxSize    :: Int
scMaxDepth   :: Maybe Int
\end{code}
%
\noindent
The ``\ttp{sc}'' prefix stands for \emph{S}mart\emph{C}heck and is used to
prevent name clashes with QuickCheck's flags.  \ttp{scMaxFailure} controls the
number of values generated by the reduction algorithm for each sub-value
analyzed.  \ttp{scMaxSize} controls the maximum size of values generated to
replace sub-values by the reduction algorithm.  Thus, new sub-values must be
strictly smaller than the minimum of \ttp{scMaxSize} and the size of the
sub-value being replaced.  (The size of the value is with respect to the
\ttp{size} method in the \ttp{SubTypes} class.)  Finally, \ttp{scMaxDepth}
determines the maximum depth in the counterexample the reduction algorithm
should analyze.  A value of \ttp{Nothing} means that the counterexample should
be exhaustively reduced, as the algorithm is presented in
Figure~\ref{fig:reduction}.  For example, the depth of a list is determined by
its length and the depth of the binary tree defined in Section~\ref{sec:reduct}
is determined by the function \ttp{depth}:
%
\begin{code}
depth L         = 0
depth (B t0 t1) =
  1 + max (depth t0) (depth t1)
\end{code}
%
\noindent
Of the flags, \ttp{scMaxDepth} is the most important for controlling
efficiency, particularly for large product types with significant ``fan out''.
The number of sub-values of a value product type can grow exponentially with
respect to the depth.  Furthermore, note that as the reduction algorithm
descends further, there is less chance to reduce the size of the value overall,
since smaller and smaller sub-values are replaced.

@@

\begin{figure}
  \begin{code}
extrapolate :: SubTypes a
       => a -> (a -> Property) -> IO [Idx]
extrapolate cex prop = extrapolate' 1 []
  where
  extrapolate' idx idxs
    | or (map (subTree cex idx) idxs) =
      extrapolate' (idx+1) idxs
  extrapolate' idx idxs
    | otherwise                       =
      case index cex idx of
        Nothing -> return idxs
        Just v  -> do
          vs <- newVals v scMaxSize scMaxSuccess
          if allFail cex idx vs prop
            then extrapolate' (idx+1) (idx:idxs)
            else extrapolate' (idx+1) idxs

allFail :: SubTypes a
     => a -> Idx -> [SubVal]
     -> (a -> Property) -> Bool
allFail cex idx vs prop =
  length res >= scMinExtrap && and res
  where
  res  = catMaybes (map go vs)
  go   = fail prop . replace cex idx
  \end{code}
  \caption{Universal sub-value generation algorithm.}
  \label{fig:universal}
\end{figure}

@@

@@

\section{Implementation}

@@

\todo{GADTs? don't work}

@@

\todo{xmonad}

@@

>>>>

data SubVal =
  forall a. SubTypes a => SubVal a

class Arbitrary a => SubTypes a where
  size    :: a -> Size
  index   :: a -> Idx -> Maybe SubVal
  replace :: a -> Idx -> SubVal -> a
  constr  :: a -> String
  constrs :: a -> [String]
\end{code}
%
\noindent
\ttp{SubTypes} has five methods.  The \ttp{size} method returns the size of a
value, \ttp{index} returns a sub-value at an breadth-first index in a value,
\ttp{replace} replaces a sub-value at a particular focus, \ttp{constr} returns
the constructor tagging the value, and \ttp{constrs} returns a list of unique
identifiers for the constructors associated with each variant of the type.

For example, consider a binary tree type:
%
\begin{code}
data Tree = L | B Tree Tree
\end{code}
%
\noindent
and the value \ttp{tree}, labeled with the indexes of its sub-values
%
\begin{code}
tree = B\sub{0} (B\sub{1} L\sub{3}
             (B\sub{4} L\sub{6} L\sub{8}))
          (B\sub{2} L\sub{5} L\sub{7})
\end{code}
%
\noindent
Here are example applications of \ttp{SubTypes} methods; in the following, we
show the indexes with respect to the \ttp{tree} value:
%
\begin{code}
size tree = 9

index tree 4 = (Just . SubVal) (B\sub{4} L\sub{6} L\sub{8})
index tree 12 = Nothing

replace tree 2 (SubVal L) =
  B\sub{0} (B\sub{1} L\sub{3}
        (B\sub{4} L\sub{6} L\sub{8}))
     L

constr  tree = ["B"]
constrs tree = ["B", "L"]
constrs L    = ["B", "L"]
\end{code}
%
\noindent
We will sometimes refer to the unwrapped value returned by \ttp{index a i} as
the \emph{$i$th sub-value of $a$}, so for example, \ttp{B\sub{4} L\sub{6}
  L\sub{8}} is the 4th sub-value of \ttp{tree}.  An invariant of \ttp{index} is
that for any value \ttp{a}, consider the least $i \geq 0$ such that \ttp{index a
  i == Nothing}.  For all $0 \leq j < i$, \ttp{index a j /= Nothing}.  We use
this invariant as a termination case in recursive algorithms over the sub-values
of a value.

The method \ttp{replace} returns the original value for an out-of-bounds index
or an incorrectly-typed replacement value.  The \ttp{SubVal} type is an
existential data type, used as a generic container for sub-values from a
counterexample.

Finally, while \ttp{constr} returns a string representation of the constructor
associated with the specific variant passed to it, \ttp{constrs} ignores the
value and returns the same list of constructors for any value from the type.
(In our implementation, \ttp{constrs} depends on GHC Generics~\cite{generics},
which we describe in Section~\ref{sec:implementation}.  For simplicity, we omit
here Generics-specific super-class constraints on the \ttp{SubTypes} class.)

\begin{figure}[ht]
  \lstset{caption={}
%         ,xleftmargin=5ex,firstnumber=0,numbers=left
         }
  \begin{code}
getSize :: SubVal -> Size
getSize (SubVal a) = size a

newVals :: SubVal -> Size -> Int -> IO [SubVal]
newVals (SubVal a) sz tries =
  replicateM tries s
  where
  s  = liftM SubVal (sizedArbitrary sz a)

reduce :: SubTypes a
       => a -> (a -> Property) -> IO a
reduce cex prop = reduce' 1
  where
  reduce' idx =
    case index cex idx of
      Nothing -> return cex
      Just v  -> do
        vs <- newVals v (getSize v) scMaxReduce
        case test cex idx vs prop of
          Nothing   -> reduce' (idx+1)
          Just cex' -> reduce cex' prop

@@

\todo{explain getSize}

\ttp{sizedArbitrary} returns a value inside the \ttp{IO} monad since it depends
on a random seed for generation (we could pass in the generator seed, but for
simplicity, we assume one is generated locally).  The constant \ttp{scMaxReduce} is
the maximum number of tries to discover a new counterexample by replacing
\ttp{v} in \ttp{cex} and testing it.  The result of \ttp{pass prop cex'} for
%
\begin{code}
pass :: (a -> Property) -> a -> Bool
\end{code}
%
\noindent
holds if \ttp{cex'} satisfies the property \ttp{prop}.  If no failure is found,
we move to the next sub-value of \ttp{cex} and continue.  However, if a new
smaller counterexample \ttp{cex'} is found, we start a new breadth-first
traversal of \ttp{cex'}, attempting to shrink it further.

@@

  test' v idx = do
    vs <- newVals v (getSize v) scMaxReduce
    case test cex idx vs prop of
      Nothing   -> reduce' (idx+1)
      Just cex' -> reduce0 cex' prop

@@

SmartCheck's implementation contains flags to allow the user to customize its
behavior.  Three flags that are relevant to the reduction algorithm are the
following:
%
\begin{code}
scMaxReduce :: Int
scMaxSize   :: Int
scMaxDepth  :: Maybe Int
\end{code}
%
\noindent
The \ttp{scMaxReduce} flag controls the number of values generated by the
reduction algorithm for each sub-value analyzed.  \ttp{scMaxSize} controls the
maximum size of values generated to replace sub-values by the reduction
algorithm.  Thus, new sub-values must be strictly smaller than the minimum of
\ttp{scMaxSize} and the size of the sub-value being replaced.  (The size of the
value is with respect to the \ttp{size} method in the \ttp{SubTypes} class.)
Finally, \ttp{scMaxDepth} determines the maximum depth in the counterexample the
reduction algorithm should analyze.  A value of \ttp{Nothing} means that the
counterexample should be exhaustively reduced, as the algorithm is presented in
Figure~\ref{fig:reduction}.  For example, the depth of a list is determined by
its length and the depth of the binary tree defined in Section~\ref{sec:reduct}
is determined by the function \ttp{depth}:
%
\begin{code}
depth L         = 0
depth (B t0 t1) =
  1 + max (depth t0) (depth t1)
\end{code}
%
\noindent
Of the flags, \ttp{scMaxDepth} is the most important for controlling
efficiency, particularly for large product types with significant ``fan out''.
The number of sub-values of a value product type can grow exponentially with
respect to the depth.  Furthermore, note that as the reduction algorithm
descends further, there is less chance to reduce the size of the value overall,
since smaller and smaller sub-values are replaced.

@@

\begin{figure}
  \begin{code}
subTrees :: SubTypes a => a -> Idx -> [Idx] -> Bool
subTrees cex idx = any (subTree cex idx)

extrapolate :: SubTypes a
            => a -> (a -> Property) -> IO [Idx]
extrapolate cex prop = extrapolate' 1 []
  where
  extrapolate' idx idxs
    | subTrees cex idx idxs
    = extrapolate' (idx+1) idxs
  extrapolate' idx idxs =
    case index cex idx of
      Nothing -> return idxs
      Just v  -> do
        vs <- newVals v scMaxSize scMaxExtrap
        extrapolate' (idx+1)
          (if allFail cex idx vs prop
             then idx:idxs else idxs)

allFail :: SubTypes a
        => a -> Idx -> [SubVal]
        -> (a -> Property) -> Bool
allFail cex idx vs prop =
  length res >= scMinExtrap && and res
  where
  res  = mapMaybe go vs
  go   = fail prop . replace cex idx
  \end{code}
  \caption{Universal sub-value generation algorithm.}
  \label{fig:universal}
\end{figure}

@@

\begin{figure}
  \begin{code}
subConstr :: SubVal -> String
subConstr (SubVal a) = constr a

subConstrs :: SubVal -> [String]
subConstrs (SubVal a) = constrs a

sumTest :: SubTypes a
          => a -> (a -> Property) -> [Idx] -> IO [Idx]
sumTest cex prop exIdxs = sumTest' 1 []
  where
  sumTest' idx idxs
    | subTrees cex idx (exIdxs ++ idxs)
    = sumTest' (idx+1) idxs
  sumTest' idx idxs =
    case index cex idx of
      Nothing -> return idxs
      Just v  -> do
        vs <- newVals v scMaxSize scConstrMax
        sumTest' (idx+1)
          (if constrFail cex idx vs prop
                (subConstr v) (subConstrs v)
             then idx:idxs else idxs)

constrFail :: SubTypes a => a -> Idx -> [SubVal]
  -> (a -> Property) -> String -> [String] -> Bool
constrFail cex idx vs prop con allCons =
  constrFail' [con] vs
  where
  constrFail' cons vs'
    | length cons == length allCons
    = True
    | null vs'
    = False
    | go v == Just True
    = constrFail' cs (tail vs')
    | otherwise
    = constrFail' cons (tail vs')

    where
    v  = head vs'
    c' = subConstr v
    cs = if c' `elem` allCons
           then cons else c':cons

  go = fail prop . replace cex idx
  \end{code}
  \caption{Universal sub-value generation algorithm.}
  \label{fig:constrs}
\end{figure}

The algorithm

@@

\section{Implementation}\label{sec:implementation}

@@

\todo{GADTs? don't work}
\todo{add lazy testing---proof value is unused}

@@

\todo{xmonad}
\todo{use SC to test itself?}

@@

\todo{future work: higher-kinded data, functions}
